---
title: Processing the Cao Perturb-seq dataset
author: Aaron Lun
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Downloading the data

We obtain a Perturb-seq dataset by [Cao and Nan](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE273677).
This uses the standard 10X layout:

```{r}
dir.create("raw_data", showWarnings=FALSE)
download_cache <- function(base.url, fname) {
    path <- file.path("raw_data", fname)
    if (!file.exists(path)) {
        download.file(paste0(base.url, URLencode(fname, reserved=TRUE)), path)
    }
    path
}

url.template <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=%s&format=file&file="
samples <- c(
    GSM8434996="GWAS-rep1",
    GSM8434997="GWAS-rep2",
    GSM8434998="GWAS-rep3",
    GSM8434999="GWAS-rep4",
    GSM8435000="GWAS-rep5",
    GSM8435001="GWAS-rep6",
    GSM8435002="GWAS-rep7",
    GSM8435003="RQC-rep1",
    GSM8435004="RQC-rep2"
)

collected <- list()
library(SingleCellExperiment)
for (n in names(samples)) {
    base.url <- sprintf(url.template, n)
    prefix <- paste0(n, "_", samples[[n]])
    barcode.path <- download_cache(base.url, paste0(prefix, "_barcodes.tsv.gz"))
    features.path <- download_cache(base.url, paste0(prefix, "_features.tsv.gz"))
    matrix.path <- download_cache(base.url, paste0(prefix, "_matrix.mtx.gz"))
}
```

# Loading the counts

Then we just load everything in with `r Biocpkg("DropletUtils")`.
First we'll do the GWAS experiment:

```{r}
library(DropletUtils)
gwas.samples <- samples[startsWith(samples, "GWAS")]
gwas.sce <- read10xCounts(
    file.path("raw_data", paste0(names(gwas.samples), "_", gwas.samples, "_")),
    type="prefix",
    mtx.class="SVT_SparseMatrix"
)

# Simplify the samples.
gwas.sce$Sample <- sub("_.*", "", basename(gwas.sce$Sample))

# Split by feature type, but no need for the actual column anymore.
gwas.type <- rowData(gwas.sce)$Type
rowData(gwas.sce)$Type <- NULL
gwas.sce <- splitAltExps(gwas.sce, gwas.type)

# Simplify the symbol designations for the guides.
rowData(altExp(gwas.sce))$Symbol <- sub("-[0-9]+$", "", rowData(altExp(gwas.sce))$Symbol)
gwas.sce
```

```{r, echo=FALSE}
gc(full=TRUE)
```

And we do the same for the RQC samples:

```{r}
library(DropletUtils)
rqc.samples <- samples[startsWith(samples, "RQC")]
rqc.sce <- read10xCounts(
    file.path("raw_data", paste0(names(rqc.samples), "_", rqc.samples, "_")),
    type="prefix",
    mtx.class="SVT_SparseMatrix"
)

rqc.sce$Sample <- sub("_.*", "", basename(rqc.sce$Sample))

rqc.type <- rowData(rqc.sce)$Type
rowData(rqc.sce)$Type <- NULL
rqc.sce <- splitAltExps(rqc.sce, rqc.type)

rowData(altExp(rqc.sce))$Symbol <- sub("-[0-9]+$", "", rowData(altExp(rqc.sce))$Symbol)
rqc.sce
```

```{r, echo=FALSE}
gc(full=TRUE)
```

# Saving to file

We apply some polish to optimize the disk space:

```{r}
library(scRNAseq)
gwas.sce <- polishDataset(gwas.sce)
gwas.sce
rqc.sce <- polishDataset(rqc.sce)
rqc.sce
```

We now save all of the relevant components of `sce` to file for upload.

```{r}
output.dir <- "2025-10-10_output"
unlink(output.dir, recursive=TRUE)
dir.create(output.dir)

meta <- list(
    title="Perturb-seq (CRISPR-based screens with single-cell RNA-sequencing readouts) screens Endoc-betaH1 cell line targeting 61 T2D coding variants genes and 40 RQC function genes.",
    description="Two set of genes were selected for Perturb-seq experiments. sgRNAs targeting each gene were selected according to previously published algorithm and cloned into the lentivirus construct containing sgRNA cassette available for 3’ direct capture Perturb-seq: hU6-gRNA with CR1CS1-EF1α-BFP.",
    taxonomy_id="9606",
    genome="GRCh38",
    sources=list(
        list(provider="GEO", id="GSE273677")
    ),
    maintainer_name="Aaron Lun",
    maintainer_email="infinite.monkeys.with.keyboards@gmail.com"
)

gwas.meta <- meta
gwas.meta$description <- paste0(gwas.meta$description, "\n\nThis object contains Perturb-seq data for the set of 61 genes with T2D coding variants.")
saveDataset(gwas.sce, file.path(output.dir, "gwas"), gwas.meta)

rqc.meta <- meta
rqc.meta$description <- paste0(rqc.meta$description, "\n\nThis object contains Perturb-seq data for the set of 40 RQC function genes.")
saveDataset(rqc.sce, file.path(output.dir, "rqc"), meta)
```

# Session information {-}

```{r}
sessionInfo()
```
